---
title: "Lab6_DataBetas"
author: "Brandon Nelsen, Burke Parent, Drew Hoffman, David Washburn, Tanner Ball"
date: "10/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(tidyverse)
COflights <- read_csv("https://files.osf.io/v1/resources/3z5as4pej8/providers/osfstorage/5a8ca28f57103100104584db")
```


#Overall Questions, Significance of Overall Questions:


#Team Plots & Questions:


#Team Recommended Course of Action:

#Individual Sections:

Tanner's Plot:
```{r}
# count the number of delayed flights to Denver
carrierDelay <- filter(COflights, DEST == 'DEN', ARR_DELAY >= 15, !is.na(ARR_DELAY)) %>% group_by(CARRIER) %>% summarise('total' = n())

# count the total number of planes landing in Denver in the dataset
landed <- filter(COflights, DEST == 'DEN', !is.na(ARR_TIME)) %>% group_by(CARRIER) %>% summarise('total' = n())

merged <- merge(carrierDelay, landed, by="CARRIER")

merged <- mutate(merged, probCarrierDelay = total.x/total.y)

ggplot(data = merged,aes(x = CARRIER, y = probCarrierDelay)) + geom_point() + geom_text(aes(label=signif(probCarrierDelay, digits = 3)), hjust=-.25, vjust=0) + xlab('Airline') + ylab('Probability of an Arrival Delay') + ggtitle('The Probability of an Arrival Delay in Denver by Airline')
```

David's Plot:

Findings: I used the data to determine how Christmas effected delays at DIA. I calculated the probability of a flight being delayed the week of Christmas by creating multiple subsets of the data and creating a new variable that determined if a flight was delayed. The probability of a flight being delayed Christmas week turned out to be 26.7%. The probability of a flight being delayed in december was 18.3% and the probability of any flight being delayed throughout all months was 17.1%. Hence, it is easy to see that the Christmas holiday has an effect on the chances of a flight being delayed. This is supported by the histogram which shows that the most amount of delayed flights occured on December 21st.

```{r}
#Create a variable specific to Denver

DENflights <- 
  COflights %>%
  filter(ORIGIN == 'DEN' | DEST == 'DEN')

#Delayed arrivals in Denver
#Create an operator that determines if a flight had a delayed arrival

DelayedOperator <-
  mutate(DENflights, Delayed = ARR_DELAY >= 15)

#Show only the delayed flights

DelayedDENFlights <-
    filter(DelayedOperator, Delayed == TRUE)


#How do major holidays effect arrivals?

DelayedDecemberFlights <-
  filter(DelayedDENFlights, MONTH == 12)

#Calulate the probabliity a flight was delayed on the week of Christmas (Dec.18-Dec.25)
# P(Delayed Christmas Week) = Delays Christmas Week/ Number of Flights Christmas Week

DelayedChristmasWeek <-
  filter(DelayedDecemberFlights, DAY_OF_MONTH %in% c(18,19,20,21,22,23,24,25))

FlightsChristmasWeek <-
  filter(DENflights, MONTH == 12, DAY_OF_MONTH %in% c(18,19,20,21,22,23,24,25))

nrow(DelayedChristmasWeek)/nrow(FlightsChristmasWeek)

## [1] 0.2673465

#Probability of any flight being delayed

nrow(DelayedDENFlights)/nrow(DENflights)

## [1] 0.1709883

#probability of a December flight being delayed

DecemberFlights <-
  filter(DENflights, MONTH == 12)

nrow(DelayedDecemberFlights)/nrow(DecemberFlights)

## [1] 0.1826364

#P(Delayed Christmas Week) = 26.7%, P(Any flight being Delayed) = 17.1%, P(December Flight being delayed) = 18.3%

#Create a plot that shows there are more delays Christmas Week oppose to the rest of December

ggplot(DelayedDecemberFlights, aes(x = DAY_OF_MONTH)) +
  geom_histogram(color = "black", fill = "white") +
  ylab('Number of Delayed Flights') +
  xlab('Day of the Month') +
  ggtitle('Distribution of Delayed Flights in December')
```

Burke's Plot:

Drew's Plot:

Brendan's Plot:

#Summary/Division of Labor: