---
title: "Lab6_DataBetas"
author: "Brandon Nelsen, Burke Parent, Drew Hoffman, David Washburn, Tanner Ball"
date: "10/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(tidyverse)
COflights <- read_csv("https://files.osf.io/v1/resources/3z5as4pej8/providers/osfstorage/5a8ca28f57103100104584db")
```


# Overall Questions, Significance of Overall Questions:




#Team Plots & Questions:
**What Factors Lead to On-time Arrivals in Denver?**
This question is important to airports because DIA wants to maximize the efficiency of their service.  The world runs on a schedule and being able to consistently rely on flight times being accurate helps people plan ahead with more certainty. If a diplomat has an important international meeting with a foreign consulate, they need to be able to plan out their trip ahead of time and factor in time of flight. If the flight time is unreliable, it causes a host other logistical issues for passengers.  Minimizing delays also allows the airport to send out and receive a higher volume of flights, increasing their financial bottom line. Hidden in the data are the factors that lead to on-time arrivals, and by analyzing it we may gain more insight on these influences.
```{r}
DENflights <- COflights %>% filter(DEST == 'DEN') %>% summarise('DEN' = n())
##filtering flights that are only arriving in denver
PtoDEN <- nrow(DENflights)/nrow(COflights)
##probability of a flight to denver out of total flights.
late_arrival <- DENflights %>% filter(!is.na(DENflights$ARR_DELAY >= 15)) %>% summarise('late' = n())
##creating late arrival to test probability
PLateArr <- late_arrival/DENflights
##the probability of a flight arriving late is 0.157

##testing factors that increase arrival delay.
late_departure <- DENflights %>% filter(DEP_DELAY >= 15)
PLateDep <- late_departure/DENflights
##probability of a late departure is 0.164

##taxi time total
NewDENflights <- mutate(DENflights, TaxiTime = TAXI_OUT + TAXI_IN)

#probability of an arrival delay with a weather delay vs arrival delay with no weather dealy. (0.157)*P(weather delay)
weather_delay_total <- DENflights %>% filter(WEATHER_DELAY == 1)
PWeatherDelay <- nrow(weather_delay_total)/nrow(late_arrival)

no_weather_delay <- DENflights %>% filter(WEATHER_DELAY == 0) 
PNoWeatherDelay <- nrow(no_weather_delay)/nrow(late_arrival)
##probability of late arrival and weather delay
PWeatherandLate <- PLateArr * PWeatherDelay ##0.0001
PNoWeatherLate <- PNoWeatherDelay * PLateArr ##0.149


##probability a flight is late due to a delayed takeoff
##probability a flight arrives later than 15 min 0.157
##probability a flight departs later than 15 min 0.169
PLate <- (PLateDep*PLateArr)/PLateArr
##0.0265

```


#Team Recommended Course of Action:


# Team Plots & Questions:


# Team Recommended Course of Action:
There are a multitude of factors affecting the timeliness of a flight, including weather, passenger volume, security concerns, time of year, among others. They each affect the arrival time in different magnitudes and with different frequencies. The question is important because it will show DIA what they need to focus on the most in terms of minimizing delay. For instance, if many more flights face security delays than carrier delays, then the airport would be better served to make their security screening process more efficient. By answering this question, it reveals what should be focused on the most and what the main obstacles in facilitating a timely flight are. 
# Individual Sections:

***Tanner's Plot:***

I chose to look at how often a particular airline arrives in Denver later than scheduled. I first got the total number of late flights in Denver, sorted by airline, then I got the total number of flights that landed in Denver, also sorted by airline. Then I created a new table that contains both of these numbers for each carrier, then added a new variable for the probability of a late arrival in Denver for a given airline.

```{r}
# count the number of delayed flights to Denver
carrierDelay <- filter(COflights, DEST == 'DEN', ARR_DELAY >= 15, !is.na(ARR_DELAY)) %>% group_by(CARRIER) %>% summarise('total' = n())

# count the total number of planes landing in Denver in the dataset
landed <- filter(COflights, DEST == 'DEN', !is.na(ARR_TIME)) %>% group_by(CARRIER) %>% summarise('total' = n())

# create a new table with the delay and total landed for all carriers
merged <- merge(carrierDelay, landed, by="CARRIER")
# add a new column that calculates the probability of a delayed arrival to Denver by carrier
merged <- mutate(merged, probCarrierDelay = total.x/total.y)

ggplot(data = merged,aes(x = CARRIER, y = probCarrierDelay)) + geom_point() + geom_text(aes(label=signif(probCarrierDelay, digits = 3)), hjust=-.25, vjust=0) + xlab('Airline') + ylab('Probability of an Arrival Delay') + ggtitle('The Probability of an Arrival Delay in Denver by Airline')
```

After plotting the results, we can see that the airlines with the highest likelihood of a late arrival in Denver in 2017 were B6 (JetBlue) and VX (Virgin America). These two carriers were much more likely to be late than the rest of the carriers in the dataset, so I was curious about potential reasons for this. I chose to look at the origin of these airlines' flights to Denver:

```{r}
B6origins <- filter(COflights, DEST == 'DEN', CARRIER == 'B6') %>% group_by(ORIGIN) %>% summarise('total' = n())

VXorigins <- filter(COflights, DEST == 'DEN', CARRIER == 'VX') %>% group_by(ORIGIN) %>% summarise('total' = n())

B6origins
VXorigins
```

As we can see, JetBlue flies to Denver exclusively from JFK in Queens, NY and BOS in Boston, MA, while Virgin America flies to Denver exclusively from SFO in San Francisco, CA. Potential reasons for the late arrivals, then, could be the flight time (since these are all pretty long flights), or it could be due to how busy these three airports are.

***David's Plot:***

Findings: I used the data to determine how Christmas effected delays at DIA. I calculated the probability of a flight being delayed the week of Christmas by creating multiple subsets of the data and creating a new variable that determined if a flight was delayed. The probability of a flight being delayed Christmas week turned out to be 25.4%. The probability of a flight being delayed in december was 18.1% and the probability of any flight being delayed throughout all months was 15.7%. Hence, it is easy to see that the Christmas holiday has an effect on the chances of a flight being delayed. This is supported by the histogram which shows that the most amount of delayed flights occured on December 21st.

```{r}
#Create a variable specific to Denver

DENflights <- 
  COflights %>%
  filter(DEST == 'DEN')

#Delayed arrivals in Denver
#Create an operator that determines if a flight had a delayed arrival

DelayedOperator <-
  mutate(DENflights, Delayed = ARR_DELAY >= 15)

#Show only the delayed flights

DelayedDENFlights <-
    filter(DelayedOperator, Delayed == TRUE)


#How do major holidays effect arrivals?

DelayedDecemberFlights <-
  filter(DelayedDENFlights, MONTH == 12)

#Calulate the probabliity a flight was delayed on the week of Christmas (Dec.18-Dec.25)
# P(Delayed Christmas Week) = Delays Christmas Week/ Number of Flights Christmas Week

DelayedChristmasWeek <-
  filter(DelayedDecemberFlights, DAY_OF_MONTH %in% c(18,19,20,21,22,23,24,25))

FlightsChristmasWeek <-
  filter(DENflights, MONTH == 12, DAY_OF_MONTH %in% c(18,19,20,21,22,23,24,25))

nrow(DelayedChristmasWeek)/nrow(FlightsChristmasWeek)

## [1] 0.2539916

#Probability of any flight being delayed

nrow(DelayedDENFlights)/nrow(DENflights)

## [1] 0.1565756

#probability of a December flight being delayed

DecemberFlights <-
  filter(DENflights, MONTH == 12)

nrow(DelayedDecemberFlights)/nrow(DecemberFlights)

## [1] 0.1804948

#P(Delayed Christmas Week) = 25.4%, P(Any flight being Delayed) = 15.7%, P(December Flight being delayed) = 18.1%

#Create a plot that shows there are more delays Christmas Week oppose to the rest of December

ggplot(DelayedDecemberFlights, aes(x = DAY_OF_MONTH)) +
  geom_histogram(color = "black", fill = "white") +
  ylab('Number of Delayed Arrivals') +
  xlab('Day of the Month') +
  ggtitle('Distribution of Delayed Arrivals in December')
```

***Burke's Plot:***
I found the differences in probability of some of the airports of which the flights departed from en route to DIA. I made a bar chart in order to display some of the airports that had the most delayed flights. We can clearly see that the airports with the most delayed flights are SFO and LAX.
```{r, echo=FALSE}
library(tidyverse)
library(dplyr)

Denverflights <- COflights %>% filter(DEST=='DEN' )
Denverflights <- Denverflights %>% mutate(LATE_ARR = (ARR_DELAY >= 15 ))
Denverflights <- Denverflights %>% filter(!is.na(ARR_DELAY))
Delayedflights <- Denverflights %>% filter(LATE_ARR == TRUE)

OriginFlight <- Delayedflights %>% group_by(ORIGIN) %>% summarize('total' = n())
TopOriginFlight <- OriginFlight %>% filter(total > 600)

#Top 15 Origins of Departure with Delayed Flights
ggplot(data = TopOriginFlight) + geom_col(mapping = aes(x = ORIGIN, y = total, fill = ORIGIN)) +
  xlab('Origin of Flight by Airport')+
  ylab('Total Count')+
  ggtitle('Total Delayed Flights by Origin of Departure')


```
***Drew's Plot:***

I used the data to determine the frequency of the different types of delays and how long they take based on the type. I first filtered the data by "delayed" flights that missed their arrival time by at least 15 minutes. After this, I broke down each delayed flight by the type, including carrier, weather, NAS, and security. From this breakdown, I calculated the average number of minutes that each type of delay took. From this, we can see that carrier and NAS delays are the most prevalent, but weather delays take the longest on average. I then created a pie chart to show the probablity of each delay more clearly, and which type is the most prevalent.

```{r, echo=FALSE}
library(tidyverse)
library(dplyr)
#filter by delayed flights
delayed <- filter(COflights, DEST == "DEN", ARR_DELAY >= 15, !is.na(ARR_TIME))

#filter by each type
carrier <- filter(delayed, CARRIER_DELAY >= 1) 
weather <- filter(delayed, WEATHER_DELAY >= 1)
nas <- filter(delayed, NAS_DELAY >= 1)
security <- filter(delayed, SECURITY_DELAY >= 1)

#mean delay per type
a_c <- mean(carrier$CARRIER_DELAY)
a_w <- mean(weather$WEATHER_DELAY)
a_n <- mean(nas$NAS_DELAY)
a_s <- mean(security$SECURITY_DELAY)
means <- c(a_c, a_w, a_n, a_s)

#total num of delays per type
t_c <- carrier %>% summarise(n())
t_w <- weather %>% summarise(n())
t_n <- nas %>% summarise(n())
t_s <- security %>% summarise(n())
t <- delayed %>% summarise(n())
totals <- c(17754, 1663, 18086, 89)

newframe <- data.frame(
  x = totals,
  y = means)

ggplot(data = newframe, aes(x, y)) + geom_point() + 
  ylab('Average Delay (minutes)') + 
  xlab('Amount of Delays') +
  ggtitle('Frequency and Average Duration of Delay by Type') + 
  annotate(geom="text", x=17754, y=41.3, label="Carrier", color="black") +
  annotate(geom="text", x=1663, y=57.8, label="Weather", color="black") +
  annotate(geom="text", x=18086, y=27.2, label="NAS", color="black") +
  annotate(geom="text", x=89, y=29.7, label="Security", color="black") +
  ylim(0,60)

lbls <- c("Carrier", "Weather", "NAS", "Security")
pie(totals, labels = lbls, main="Delays by Type")
```

***Brandon's Plot:***

#Summary/Division of Labor: